---
title: "generic_preprocessing"
format: html
editor: source
---

# Setup & Raw Data Investigation

This document provides a *generic preprocessing pipeline* using the HMGU
reference method for both positive and negative modes. and is set up for Human
endosome.

## Load Required Packages and functions

```{r, message=FALSE, warning=FALSE}
base_path <- file.path("1_preprocessing")

lab <- "afekta"
lab_path <- file.path(base_path, lab)

study_group <- "HE" #example HE / FE / Others
study_path <- file.path(base_path, lab, study_group)

source(file.path(base_path, "setup.R"))
```

In case one or more of the above libraries are not available (and hence fail to
load), they can be installed with the R command `BiocManager::install()` (e.g.
`BiocManager::install("alabaster.base")` to install the *alabaster.base* R
package). Note that this requires the installation of the *BiocManager* R
package (which can be installed with `install.packages("BiocManager")`).

## Load metadata and raw file

```{r load-lcms-data, warning=FALSE, message=FALSE}
# Load pos data
seq_pos <- read_xlsx(file.path(study_path, "seq_pos.xlsx"), col_names = TRUE) |>
  as.data.frame()
seq_pos$polarity <- "pos"
# load neg data
seq_neg <- read_xlsx(file.path(study_path, "seq_neg.xlsx"), col_names = TRUE) |>
  as.data.frame()
seq_neg$polarity <- "neg"
seq <- rbind(seq_pos, seq_neg)

seq$filename <- paste0(seq$`Data File`, ".mzML")
#filtering the files because i have a subset of it- remove this later
seq <- seq[seq$filename %in% list.files(file.path(study_path, "mzml/")), ]
mse <- readMsExperiment(
  paste0(file.path(study_path, "mzml"), "/", seq$filename),
  seq
)

sampleData(mse)
```

Below we set up the parallel processing, which will allow to make the
pre-processing MUCH more efficient. please adjust to the number of cores that
your computer can handle (mine has 4 so I use 2 or 3 for data analysis)

This number will then be the same as the one you input in the `chunkSize =`
parameter in the preprocessing steps.

```{r}
cores_nb <- 2 # Adjust the number of cores based on your machine
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(cores_nb))
} else {
    register(SnowParam(cores_nb))
}
```

## Overview and Quality

Split pos and neg for this part.

```{r}
mse_pos <- mse[sampleData(mse)$polarity == "pos"]
mse_neg <- mse[sampleData(mse)$polarity == "neg"]

col_sample_type <- brewer.pal(
    n = max(3, length(unique(sampleData(mse_pos)$sample_type))), "Set2")
names(col_sample_type)[seq_along(unique(sampleData(mse_pos)$sample_type))] <-
    unique(sampleData(mse_pos)$sample_type)
```

Suggestion to extract the BPC of all samples and plot them, all in one. BPC

```{r, eval=FALSE}
#| fig-height: 8
#| fig-width: 6
bpc_pos <- chromatogram(mse_pos, aggregationFun = "max")
bpc_neg <- chromatogram(mse_neg, aggregationFun = "max")

xl <- range(unlist(lapply(bpc_pos, rtime)), unlist(lapply(bpc_neg, rtime)))
yl <- max(unlist(lapply(bpc_pos, intensity)),
          unlist(lapply(bpc_neg, intensity)), na.rm = TRUE)
plot(3, 3, pch = NA, xlim = xl, ylim = c(-yl, yl), xlab = "rtime",
     ylab = "intensity", main = "BPC")
abline(h = 0)
grid()

for (i in seq_len(ncol(bpc_pos))) {
    points(rtime(bpc_pos[, i]), intensity(bpc_pos[, i]), type = "l",
           col = paste0(col_sample_type[bpc_pos$sample_type[i]], 80))
}
for (i in seq_len(ncol(bpc_neg))) {
    points(rtime(bpc_neg[, i]), -intensity(bpc_neg[, i]), type = "l",
           col = paste0(col_sample_type[bpc_neg$sample_type[i]], 80))
}

par(mfrow = c(2, 1))
plot(bpc_pos, col = paste0(col_sample_type[bpc_pos$sample_type], 80))
grid()
legend("topright", col = col_sample_type, lwd = 1,
       legend = names(col_sample_type))
plot(bpc_neg, col = paste0(col_sample_type[bpc_neg$sample_type], 80))
grid()
```

Observations: ...

RT ranges and spectra number per MS levels:

```{r, eval=FALSE}
#' Range
range(rtime(mse_pos))
range(rtime(mse_neg))

#' Count the number of spectra with a specific MS level per file.
print("Positive mode")
spectra(mse_pos) |>
    msLevel() |>
    split(fromFile(mse_pos)) |>
    lapply(table) |>
    do.call(what = cbind)

#' Count the number of spectra with a specific MS level per file.
print("Negative mode")
spectra(mse_neg) |>
    msLevel() |>
    split(fromFile(mse_neg)) |>
    lapply(table) |>
    do.call(what = cbind)
```

Obervations: ...

Below decide if want to filter retention time (method dependent), if yes run:

```{r}
mse <- filterRt(mse, rt = c(18, 720))
```

# Estimate peak-picking parameters on ONE file.

```{r, eval = TRUE}
x_meta <-  split(meta, meta$Mixture)[[1]]
idx_sample <- which(sampleData(mse)$sample_type == "sample")[1]
x_mse <- mse[idx_sample]
```

## Extract theoretical EICs

Using theoretical m/z for both main positive adducts and all along the
retention time.

```{r, eval = TRUE}
#| code-fold: true
#| code-summary: "Show the code"

#' Calculate m/z from the exact mass
theo_mz <- mass2mz(x_meta[, "M"], c("[M+H]+", "[M+Na]+")) |>
    as.vector()
#' Define the m/z range.
theo_mzr <- cbind(mzmin = theo_mz - MsCoreUtils::ppm(theo_mz, 20),
                  mzmax = theo_mz + MsCoreUtils::ppm(theo_mz, 20))
#' Expand by a fixed tolerance.
theo_mzr[, 1L] <- theo_mzr[, 1L] - 0.05 # increase if missing data points, decrease if too much noise.
theo_mzr[, 2L] <- theo_mzr[, 2L] + 0.05
eicsMH <- chromatogram(x_mse, mz = theo_mzr)

#' would use the ID + adduct as file name.
fData(eicsMH)$std_id <- paste0(rep(x_meta$ChEBI, 2), "-",
                               rep(c("[M+H]+", "[M+Na]+"), each = nrow(x_meta)))
fData(eicsMH)$std_name <- paste0(rep(x_meta$`ChEBI name`, 2), "-",
                               rep(c("[M+H]+", "[M+Na]+"), each = nrow(x_meta)))
rownames(eicsMH) <- fData(eicsMH)$std_id

dr <- paste0(study_path, "eics_full_rt_slices/M_ions/")
dir.create(dr, recursive = TRUE, showWarnings = FALSE)
for (i in seq_along(eicsMH)) {
    cmp <- fData(eicsMH)$std_id[i]
    cmp <- gsub(":", "_", cmp)
    png(paste0(dr, "EIC_", cmp, ".png"),
        width = 12, height = 8, units = "cm", res = 600, pointsize = 4)
    plot(eicsMH[i, ], main = fData(eicsMH)$std_name[i])
    legend("topright",
           legend = c(mzmin = format(fData(eicsMH)$mzmin[i], digits = 6),
                      mzmax = format(fData(eicsMH)$mzmax[i], digits = 6)))
    grid()
    dev.off()
}
```

We have `r length(theo_mz)` number of standard.

## Peak picking param set up

Look at the previously generated EICs to set up parameters. This should happen
only once per lab.

First quickly check a few peaks to look for peak-width:

```{r, eval=FALSE}
## !!! adjust to compounds you see interesting peaks, look for multiple !!!
plot(eicsMH["CHEBI:17485-[M+H]+"], xlim = c(20, 50)) # 10 s
plot(eicsMH["CHEBI:16828-[M+H]+"], xlim = c(80, 120)) # 10 s ## this one does not work
plot(eicsMH["CHEBI:139582-[M+Na]+"], xlim = c(18, 100)) # how large is the peak ?
```

Set up a lower and upper boundary of peak-width based on your observation.

```{r}
pw <- c(5, 15) ## define here based on the observations above
```

For ppm deviation:

Choose a compound from above and adjust the

```{r, eval=FALSE}
mz <- mz(eicsMH["CHEBI:17485-[M+H]+"]) |> as.vector() ## choose compound here

x_mse |>
    filterMsLevel(1L) |>
    filterRt(c(28, 34)) |>
    filterMzRange(mz = c(269.015, 269.025)) |> # play around with this to determinate the best range
    plot()

cst <- x_mse |>
    spectra() |>
    filterMsLevel(msLevel = 1L) |>
    filterRt(rt = c(28.5, 33.5)) |> # be restrictive
    filterMzRange(mz =  c(269.015, 269.025)) # set it up as above.

#' Show the number of peaks per m/z filtered spectra
lengths(cst) ### the output here should look a bit like "1 1 1 1 1 "
```

```{r, eval=FALSE}
#' Calculate the difference in m/z values between scans
mz_diff <- cst |>
    mz() |>
    unlist() |>
    diff() |>
    abs()

#' Express differences in ppm
range(mz_diff * 1e6 / mean(unlist(mz(cst))))
```

Take a number a bit above the output from above. Also depends on the machine.
We try not to go below 10ppm.

```{r}
ppmval <- 20 ## input your number here
```

# Peak picking on ALL samples

```{r}
cwp <- CentWaveParam(peakwidth = pw,
                     ppm = ppmval,
                     integrate = 2,
                     snthresh = 7, ## lower if your dataset is too noisy.
                     noise = 100)  ## increase depending on what you estimate you minimum intensity for real signal to be
mse <- findChromPeaks(mse, cwp,
                          msLevel = 1L, chunkSize = cores_nb)

mdpp <- MergeNeighboringPeaksParam(expandRt = max(pw)/2,
                                   expandMz = 0.0015,
                                   minProp = 0.75)
mse <- refineChromPeaks(mse, mdpp, chunkSize = cores_nb)

chromPeakData(mse)$merged |>
                         table()
```

# Alignment using NAPS

```{r naps-part1}
#' loading info from NAPS
naps_info <- read_xlsx(file.path(base_path, "NAPS_info.xlsx"),
                       .name_repair = "minimal") |>
    as.data.frame()

#' calculate m/z
naps_info$mz_pos <- mass2mz(naps_info$`exact mass`, adduct = "[M+H]+")[, 1]

#' subset positive polarity
naps_pos <- mse[sampleData(mse)[, "sample_type"] == "NAPS" &
                sampleData(mse)[, "polarity"] == "pos"]

#' Match chromatographic peaks against NAPS
cpks <- chromPeaks(naps_pos)
naps_pos_match <- matchValues(cpks, naps_info, mzColname = c("mz", "mz_pos"),
                              MzParam(tolerance = 0, ppm = 10))
naps_pos_match <- naps_pos_match[whichQuery(naps_pos_match)]
md <- matchedData(naps_pos_match, c("mz", "rt", "maxo", "sample",
                                    "ppm_error", "target_Name")) |>
    as.data.frame()


pos_rt_matrix <- naps_rt_matrix(md, naps_info)
colnames(pos_rt_matrix) <- paste0("NAPS_POS_",
                                  seq_len(ncol(pos_rt_matrix)))
```

```{r naps-part2}
#' Define m/z for the expected ions of the NAPS
naps_info$mz_neg <- mass2mz(naps_info$`exact mass`, adduct = "[M+CHO2]-")[, 1]

#' Negative polarity
naps_neg <- mse[sampleData(mse)[, "sample_type"] == "NAPS" &
                sampleData(mse)[, "polarity"] == "neg"]
cpks <- chromPeaks(naps_neg)
naps_neg_match <- matchValues(cpks, naps_info, mzColname = c("mz", "mz_neg"),
                              MzParam(tolerance = 0, ppm = 10))
naps_neg_match <- naps_neg_match[whichQuery(naps_neg_match)]
md <- matchedData(naps_neg_match, c("mz", "rt", "maxo", "sample",
                                    "ppm_error", "target_Name")) |>
    as.data.frame()

neg_rt_matrix <- naps_rt_matrix(md, naps_info)
colnames(neg_rt_matrix) <- paste0("NAPS_NEG_", seq_len(ncol(neg_rt_matrix)))
matchedData(naps_neg_match, c("rt", "maxo", "sample", "target_Name"))
```

```{r naps-combine}
rt_matrix <- cbind(pos_rt_matrix, neg_rt_matrix)
## export table
write.csv(
    cbind(naps_info, rt_matrix),
    file = file.path(study_path, "naps.csv"),
    row.names = TRUE
)
```

```{r rti-setup}
idx <- which(sampleData(mse)$Sample.Name == "NAPS")
idx_sample <- which(sampleData(mse)$sample_type == "sample")

for (i in idx_sample) {
    idx_naps_bef <- get_closest_index(i, idx, "previous")
    sampleData(mse)[i, "NAPS_before"] <-
        colnames(rt_matrix)[which(idx == idx_naps_bef)]
    idx_naps_aft <- get_closest_index(i, idx, "next")
    sampleData(mse)[i, "NAPS_after"] <-
        colnames(rt_matrix)[which(idx == idx_naps_aft)]
}
```

Alignment based on the table above

```{r align}
pgp <- PeakGroupsParam(
    minFraction = 0.75,
    extraPeaks = 50,
    span = 0.5,
    subset = which(sampleData(mse)$sample_type == "NAPS"),
    subsetAdjust = "average",
    peakGroupsMatrix = rt_matrix)
mse <- adjustRtime(mse, param = pgp)


plotAdjustedRtime(
    mse, col = paste0(col_sample_type[sampleData(mse)$sample_type], 60))

mse <- applyAdjustedRtime(mse)
```

## export object for library building

```{r, export}
## export object needed for library building
mse <- mse[sampleData(mse)$sample_type == "sample"]

if (length(unique(sampleData(mse)$ms_level)) > 1) {
    mse2 <- mse[sampleData(mse)$ms_level == 2]
    mse <- mse[sampleData(mse)$ms_level == 1]
    fl <- file.path(study_path, "mse2")
    if (file.exists(fl))
        unlink(fl, recursive = TRUE)
    saveMsObject(mse2, AlabasterParam(path = fl))
    fl <- file.path(study_path, "mse")
    if (file.exists(fl))
        unlink(fl, recursive = TRUE)
    saveMsObject(mse, AlabasterParam(path = fl))
} else {
    fl <- file.path(study_path, "mse")
    if (file.exists(fl))
        unlink(fl, recursive = TRUE)
    saveMsObject(mse, AlabasterParam(path = fl))
}
```
