---
title: "chr_genderation_comparison"
format: html
execute:
  dir: project
---

### TODOS 

- FIX PCA plot 
- Introduce statistical test for the differences observed.
- fix heatmap for detected ratio similarity.
- Download library on cluster.
- fi xthe eval conditioon

# Introduction

In this analysis, I aim to create **Chromatograms objects** from the
preprocessed LC-MS data, primarily using the functionality provided by the
**Chromatograms package**. The main goal is to compare full and extracted data 
across different labs to understand how LC and MS setups influence the results.

The comparison will proceed in multiple steps:

1. **Full Data Analysis**:  

   - Evaluate basic metrics such as BPC, TIC, and median intensity across
     retention time (RT) bins.  
   - Compute basic similarity metrics to assess data consistency across labs.  
   
1. **Detected Signal Analysis**:  

   - Focus on signals detected by peak picking algorithms, excluding baseline
     noise.  
   - Recompute the same metrics and similarity measures as in the full data
     analysis.  
   - This step aims to highlight differences in sensitivity and peak detection
     efficiency across labs. We hope to get more clear information on setup 
     specificity.

2. **Annotated Signal Analysis**:   

   - Use lab-provided annotations to extract only "annotated" signals.  
   - Recompute the same metrics and similarity measures.  
   - Examine EICs of annotated peaks and calculate a variety of metrics for 
     inter-lab comparison.  
   - Save these results for later use in a separate peak similarity analysis
     workflow, which will require evaluating multiple aspects of the data.  

Currently, only harmonized method data is available, meaning the samples were
run using the same column and protocol. Therefore, most observed differences 
are expected to originate from variations in MS setups.  

For now, this analysis focuses solely on the **human endosome mixture**, but
the workflow is designed to be extendable. A separate R script,
`rt_analysis_function.R`, contains custom functions used in this workflow.  


## Load libraries

```{r library, include=TRUE, warning=FALSE, message=FALSE}
library(Chromatograms)
library(Spectra)
library(xcms)
library(MsIO)
library(MsExperiment)
library(vioplot)
library(dtw)
library(dtwclust)
library(proxy) # Added for efficient DTW calculation
library(pracma) ## easy find peak function, switch to MsCoreUtils
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggbeeswarm)

source("rt_analysis_function.R")
```

```{r, cores-cluster}
# cluster parallel setting
bpparam <- MulticoreParam(workers = parallel::detectCores() - 1)
```

# Full data analysis

Below I load the preprocessed data from each lab.

```{r load-full-data}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/sp_full.RData")) {
  sp_afekta <- load_data(lab = "afekta", study_group = "HE")
  sp_hmgu <- load_data(lab = "hmgu", study_group = "HE")
  sp_icl <- load_data(lab = "icl", study_group = "HE")
  sp_cembio <- load_data(lab = "cembio", study_group = "HE")

  sp_full <- c(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
  save(sp_full, file = "object/sp_full.RData")
  rm(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
} else {
  load("object/sp_full.RData")
}
```

If we summarise the number of MS1 spectra per lab: 

```{r nb-spectra-full}
table(sp_full$lab) 
```

Below we summarize the rtime range per lab (in seconds). 
They should be fairly similar for the harmonized method as we agreed on a 
range during preprocessing.

```{r rt-range-full}
lapply(split(rtime(sp_full), sp_full$lab), range)
```

Insert table with experimental info later below: 

[table]

## Basic metrics

### Base Peak Chromatogram (BPC) area

```{r load-bpc-full}
#| code-fold: true
#| code-summary: "Show the code"
if(!file.exists("object/bpc_full.RData")) {
  bpc_full <- Chromatograms(sp_full, summarize.method = "max",
                       factorize.by= "dataOrigin", 
                       spectraVariables = c("mixture", "lab")) 
  bpc_full <- setBackend(bpc_full, ChromBackendMemory())
  save(file = "object/bpc_full.RData", bpc_full)
} else {
  load("object/bpc_full.RData")
}
```

Below we compute the BPC area per mixture per lab and summarize across mixtures
by summing the intensity values.

Inference from Full BPC: 
The BPC area on full data represents the total intensity of the most dominant
ion at every scan, including baseline and chemical noise. Differences between 
labs here can indicate variations in overall instrument sensitivity or, more 
likely, differences in the level of background noise. A lab with a 
significantly higher BPC may have a "noisier" baseline with more 
intense background ions.

```{r sum-bpc-full, eval=TRUE}
# Create a data frame for plotting [data from cite: 59]
plot_data <- data.frame(
  Lab = bpc_full$lab,
  Mixture = bpc_full$mixture,
  BPC_Area = vapply(intensity(bpc_full), sum, numeric(1), na.rm = TRUE)
)

# Create an advanced plot
ggplot(plot_data, aes(x = Lab, y = log2(BPC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Full Data BPC Across Laboratories (POS AND NEG)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "Log2(BPC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


Note: this above is not normalized, we could by TIC but we already compute 
the ratio below and it is probably the best computation anyway. 

Next compute CV per mixture across lab:
Here normalize by standard deviation divided by mean. we are hoping to get some
information about the agreement of the BPC signal across lab.

```{r cv-full ,eval=TRUE}
sbpc <- vapply(intensity(bpc_full), sum, numeric(1), na.rm = TRUE) 
cv_per_mix <- tapply(sbpc, bpc_full$mixture, function(x) sd(x, na.rm =TRUE) / 
                       mean(x, na.rm =TRUE))
```


### Total Ion Current (TIC) area

```{r load-tic-full}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/tic_full.RData")) {
  tic_full <- Chromatograms(sp_full, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab")) 
  tic_full <- setBackend(tic_full, ChromBackendMemory())
  save(file = "object/tic_full.RData", tic_full)
} else {
  load("object/tic_full.RData")
}
```

Inference from Full TIC
The TIC area on full data sums all ion intensities in every scan. This is a 
measure of the total ion flux, or the total amount of "signal" (including all 
noise, background, and analytes) detected by the instrument. Large differences 
(e.g., 'icl' being much higher) suggest a significantly higher overall
signal response, which could be due to higher background, different tuning,
or higher sensitivity.

```{r int-tic-full, eval=TRUE}
stic <- vapply(intensity(tic_full), sum, numeric(1), na.rm = TRUE) # Need to define stic
plot_data_tic <- data.frame(
  Lab = tic_full$lab,
  Mixture = tic_full$mixture,
  TIC_Area = stic
)

# Create an advanced plot for TIC
ggplot(plot_data_tic, aes(x = Lab, y = log2(TIC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Full Data TIC Across Laboratories (POS AND NEG)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "Log2(TIC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Note: again this is not normalized. 

### ratio BPC / TIC

This metric, when applied to full data, is a powerful proxy for data 
complexity and baseline noise. A low ratio (e.g., 0.1-0.2) indicates a 
"messy" signal, where the total intensity (TIC) is spread across many 
different m/z values, and no single ion dominates. This is typical of a high,
complex chemical or electronic baseline. A higher ratio would suggest a 
"cleaner" run where, even in the baseline, a few ions dominate.

The kruskal.test result (p-value < 2.2e-16) confirms that the differences in 
this ratio between labs are statistically significant. It means at least one 
lab has a systematically different data complexity profile.

```{r ratio-full-across, eval=TRUE}
ratio_bpc_tic <- sbpc / stic
ratio_bpc_tic

# Example for BPC/TIC ratio
kruskal.test(ratio_bpc_tic ~ tic_full$lab)

plot_data_ratio <- data.frame(
  Lab = tic_full$lab,
  Mixture = tic_full$mixture,
  BPC_TIC_Ratio = ratio_bpc_tic
)

# Create an advanced plot for BPC/TIC ratio
ggplot(plot_data_ratio, aes(x = Lab, y = BPC_TIC_Ratio, fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Full Data BPC/TIC Ratio Across Laboratories (POS AND NEG)",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "BPC/TIC Ratio"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Kruskal results: Kruskal-Wallis chi-squared = 22.226, df = 3, p-value = 5.854e-05

## tic/bpc ratio along rt 

Inference from Full BPC/TIC Ratio along RT
This is a "fingerprint" of the baseline behavior.

BPC/TIC Ratio Plot: 
We are looking for differences in the shape of these profiles. A lab with a 
consistently lower profile (like 'icl') has a more complex, "messy" baseline 
across the entire gradient. A spiky profile (like 'afekta') suggests its
baseline is cleaner, and the ratio only drops in specific, complex regions.

Data Point Coverage Plot: This shows the density of data points (scans)
across the run. Ideally, all labs should be similar, as the run time was 
harmonized. Any large deviations here would indicate a problem with data 
acquisition (e.g., dropouts) or preprocessing.

```{r ratio-full, eval = TRUE, fig.width=12, fig.height=10}
# Compute median BPC/TIC ratio per RT bin
compute_bin_ratio <- function(rt, int_tic, int_bpc, n_bins = 30) {
  bins <- cut(rt, breaks = n_bins)
  tapply(seq_along(rt), bins, function(idx) {
    median(int_bpc[idx] / int_tic[idx], na.rm = TRUE)
  })
}

# Compute number of data points per RT bin
compute_bin_count <- function(rt, n_bins = 30) {
  bins <- cut(rt, breaks = n_bins)
  tapply(seq_along(rt), bins, length)
}

# Preallocate containers
ratio_all <- sapply(seq_along(tic_full), function(i) {
  rt <- rtime(tic_full[i])[[1]]
  int_tic <- intensity(tic_full[i])[[1]]
  int_bpc <- intensity(bpc_full[i])[[1]]
  
  compute_bin_ratio(rt, int_tic, int_bpc, n_bins = 30)
})

coverage_all <- sapply(seq_along(tic_full), function(i) {
  rt <- rtime(tic_full[i])[[1]]
  compute_bin_count(rt, n_bins = 30)
})

# Assign names
colnames(ratio_all) <- paste(tic_full$lab, tic_full$mixture, "pol:", tic_full$polarity, sep = "_")
colnames(coverage_all) <- colnames(ratio_all)

# --- 1. Melt the 'ratio_all' matrix ---
ratio_df <- as.data.frame(ratio_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "BPC/TIC Ratio") 

# --- 2. Melt the 'coverage_all' matrix ---
coverage_df <- as.data.frame(coverage_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "Data Point Coverage") 

# --- 3. Combine them and add lab/mixture info ---
sample_info <- data.frame(
  sample_id = colnames(ratio_all),
  lab = tic_full$lab,
  mixture = tic_full$mixture
)

plot_data <- bind_rows(ratio_df, coverage_df) %>%
  left_join(sample_info, by = "sample_id")

# --- 4. Plot ---
plot_data$metric <- factor(plot_data$metric, 
                           levels = c("BPC/TIC Ratio", "Data Point Coverage"))

ggplot(plot_data, aes(x = rt_bin, y = value, color = lab, group = sample_id)) +
  geom_line(alpha = 0.7, linewidth = 1) +
  facet_grid(metric ~ mixture, scales = "free_y") +
  labs(
    title = "Chromatographic Fingerprint Comparison (Full Data)",
    subtitle = "BPC/TIC Ratio and Data Point Coverage by Lab and Mixture",
    x = "Retention Time Bin",
    y = "Value (Ratio or Count)",
    color = "Laboratory"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12) 
  )

## for later similarity analysis i need to remove NA 
ratio_all[is.na(ratio_all)] <- 0
```


## Chromatogram  Similarity

For this we use Dynamic Time Warping (DTW), which measures similarity between 
two chromatograms (time series) while being robust to small shifts in
retention time. This makes it ideal for comparing runs where RTs may drift
slightly. A low "normalizedDistance" means the two chromatograms have a very 
similar shape.

### ratio along retention time. 

Inference from Full Ratio DTW
This is a more robust comparison. It compares the "fingerprints of complexity" 
from the previous section. If labs cluster here, it means their baseline noise
profile is fundamentally similar across the entire gradient (e.g., their noise
levels change in the same way as the gradient changes). This is a strong
indicator of similarities/differences in the MS setup's response to the mobile
phase.

```{r sim-ratio-full, eval= eval_sim, fig.width=8, fig.height=10}
# Here, ratio_all has bins as rows and samples as columns.
# We need to transpose it for proxy::dist (samples as rows).
sim_matrix_med <- as.matrix(proxy::dist(t(ratio_all), 
                                        method = "dtw_lb",
                                        window.type = "sakoechiba",
                                        window.size = 5)) # Smaller window for 30 bins

labels <- colnames(ratio_all)
rownames(sim_matrix_med) <- labels
colnames(sim_matrix_med) <- labels

## clustering
hc_med <- hclust(as.dist(sim_matrix_med))
plot(hc_med, main = "DTW clustering of full tic/bpc ratio")

heatmap(sim_matrix_med, main = "DTW similarity of full tic/bpc ratio")
```

The labels make it very hard to read. but it is very fast to run now.

```{r, rm-full}
## delete the memory hungry object 
rm(tic_full, bpc_full, sp_full)
```

# Detected signal analysis

This signal is essentially keeping the rt and mz area from peak detection. 
Now it reflects total detected signal (sensitivity of detection).

Easier to compare labs because baseline noise is excluded.
Differences now relate more to how well peaks were detected and instrument 
sensitivity rather than background noise. 

Load the results sheet below 

```{r load-detect}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/sp_full_detect.RData")) {
  bpparam <- MulticoreParam(8) # for the cluster
  sp_afekta <- detect_signal(lab = "afekta", study_group = "HE", 
                             annotated = FALSE, bpparam = bpparam) # much better.
  save(sp_afekta, file = "object/sp_afekta.RData") ## trying to save some memory space. 
  rm(sp_afekta) 
  
  sp_hmgu <- detect_signal(lab = "hmgu", study_group = "HE",
                           annotated = FALSE, bpparam = bpparam) # 
  save(sp_hmgu, file = "object/sp_hmgu.RData")
  rm(sp_hmgu) 
  
  sp_icl <- detect_signal(lab = "icl", study_group = "HE",
                          annotated = FALSE, bpparam = bpparam)
  save(sp_icl, file = "object/sp_icl.RData")
  rm(sp_icl)
  
  sp_cembio <- detect_signal(lab = "cembio", study_group = "HE",
                             annotated = FALSE, bpparam = bpparam)
  save(sp_cembio, file = "object/sp_cembio.RData")
  
  
  load("object/sp_hmgu.RData")
  load("object/sp_afekta.RData")
  load("object/sp_icl.RData")
  
  sp_full_detect <- c(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
  save(sp_full_detect, file = "object/sp_full_detect.RData")
  rm(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
} else {
  load("object/sp_full_detect.RData")
}
```

Summarize the number of MS1 spectra per lab:

```{r nb-spectra-detect}
table(sp_full_detect$lab) ## compare with sp_full 
```

Summarize the rtime range per lab (in seconds):

```{r rt-range-detect}
lapply(split(rtime(sp_full_detect), sp_full_detect$lab) , range)
```

## Basic metrics 

Now baseline noise and spurious fluctuations are largely removed.

### Base Peak Chromatogram (BPC) area

We generate the BPC from the detected signal.

```{r load-bpc-detect}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/bpc_detect.RData")) {
  bpc_detect <- Chromatograms(sp_full_detect, summarize.method = "max",
                       factorize.by= "dataOrigin", 
                       spectraVariables = c("lab", "mixture")) 
  bpc_detect <- setBackend(bpc_detect, ChromBackendMemory())
  save(file = "object/bpc_detect.RData", bpc_detect)
} else {
  load("object/bpc_detect.RData")
}
```

We summarize the BPC area per mixture per lab by summing the intensity values.

Inference from Detected BPC
This is a much more meaningful metric than the "full" BPC. It represents 
the sum of the apex intensities of all detected peaks. This is a direct proxy 
for instrument sensitivity and peak detection efficiency. A lab with a higher 
detected BPC is either detecting more peaks or detecting the 
same peaks at a higher intensity. The CV per mixture  quantifies the inter-lab 
variability in this sensitivity.

```{r sum-bpc-detect}
sbpc <- vapply(intensity(bpc_detect), sum, numeric(1), na.rm = TRUE)

plot_data_bpc_det <- data.frame(
  Lab = bpc_detect$lab,
  Mixture = bpc_detect$mixture,
  BPC_Area = sbpc
)

ggplot(plot_data_bpc_det, aes(x = Lab, y = log2(BPC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Detected BPC Area Across Laboratories",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "Log2(Detected BPC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Next, we compute the coefficient of variation (CV) per mixture across labs.

```{r cv-detect}
cv_per_mix_det <- tapply(sbpc, bpc_detect$mixture, function(x) sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
plot(cv_per_mix_det, main = "CV of Detected BPC per Mixture")
```

### Total Ion Current (TIC) area

```{r load-tic-detect}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/tic_detect.RData")) {
  tic_detect <- Chromatograms(sp_full_detect, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab")) 
  tic_detect <- setBackend(tic_detect, ChromBackendMemory())
  save(file = "object/tic_detect.RData", tic_detect)
} else {
  load("object/tic_detect.RData")
}
```

We summarize the TIC area per mixture per lab by summing the intensity values.

Inference from Detected TIC
This represents the sum of the areas of all detected peaks. It is the
best measure of total quantifiable signal. Unlike the "full" TIC, this 
excludes the baseline. Labs with higher detected TIC are capturing more total 
analyte signal, again pointing to higher sensitivity or better peak detection.

```{r int-tic-detect}
stic <- vapply(intensity(tic_detect), sum, numeric(1), na.rm = TRUE)

plot_data_tic_det <- data.frame(
  Lab = tic_detect$lab,
  Mixture = tic_detect$mixture,
  TIC_Area = stic
)

ggplot(plot_data_tic_det, aes(x = Lab, y = log2(TIC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Detected TIC Area Across Laboratories",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "Log2(Detected TIC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### ratio BPC / TIC

ow that we should have gotten rid of baseline and noise signal, the ratio 
should be properly describing the complexity of the detected signal.

Inference from Detected BPC/TIC Ratio:

This measures the **complexity of the detected features**.

A ratio closer to 1 would imply that the detected peaks 
are "pure" (BPC $\approx$ TIC).

A low ratio (like we see) means that even
within the detected peak boundaries, the signal is spread across many ions. 
This could be due to co-elution, fragmentation, adducts, or the peak-picking 
algorithm including noisy parts of the peak.

Comparing labs, afekta seems to have a lower ratio, suggesting their detected 
peaks are more complex(more co-elution/fragments) than other labs.

```{r ratio-detect-across}
ratio_bpc_tic <- sbpc / stic

plot_data_ratio_det <- data.frame(
  Lab = tic_detect$lab,
  Mixture = tic_detect$mixture,
  BPC_TIC_Ratio = ratio_bpc_tic
)

ggplot(plot_data_ratio_det, aes(x = Lab, y = BPC_TIC_Ratio, fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Detected Data BPC/TIC Ratio Across Laboratories",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "BPC/TIC Ratio (Detected Signal)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Tic/Bpc ratio along rt 

Gives us some kind of chromatographic fingerprint per lab. 
Instead of baseline shape, you’re summarizing where peaks are detected along 
the gradient.


Inference from Detected BPC/TIC Ratio along RT

This is a "fingerprint of detected signal."

BPC/TIC Ratio Plot: This shows where in the gradient the detected signal is 
"clean" (high ratio) or "complex" (low ratio). All labs show a similar
pattern (e.g., a drop in complexity in the middle bins), but at different
levels. 'icl' consistently shows the most complex (lowest ratio) signal,
while 'afekta' shows the cleanest (highest ratio).

Data Point Coverage Plot: This is now a very important plot. It is no longer 
uniform. It shows where in the gradient each lab detected signal. This is a 
powerful visualization of peak density. All labs seem to detect most of their 
signal in the first ~15 bins. 'icl' appears to detect signal across more
bins than others.

```{r ratio-detect, eval = FALSE, fig.width=12, fig.height=9}
# Preallocate containers
ratio_all <- sapply(seq_along(tic_detect), function(i) {
  rt <- rtime(tic_detect[i])[[1]]
  int_tic <- intensity(tic_detect[i])[[1]]
  int_bpc <- intensity(bpc_detect[i])[[1]]
  
  compute_bin_ratio(rt, int_tic, int_bpc, n_bins = 30)
})

coverage_all <- sapply(seq_along(tic_detect), function(i) {
  rt <- rtime(tic_detect[i])[[1]]
  compute_bin_count(rt, n_bins = 30)
})

# Assign names
colnames(ratio_all) <- paste(tic_detect$lab, tic_detect$mixture, "pol:", tic_detect$polarity, sep = "_")
colnames(coverage_all) <- colnames(ratio_all)

# --- 1. Melt the 'ratio_all' matrix ---
ratio_df <- as.data.frame(ratio_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "BPC/TIC Ratio") 

# --- 2. Melt the 'coverage_all' matrix ---
coverage_df <- as.data.frame(coverage_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "Data Point Coverage") 

# --- 3. Combine them and add lab/mixture info ---
sample_info <- data.frame(
  sample_id = colnames(ratio_all),
  lab = tic_detect$lab,
  mixture = tic_detect$mixture
)

plot_data <- bind_rows(ratio_df, coverage_df) %>%
  left_join(sample_info, by = "sample_id")

# --- 4. Plot ---
plot_data$metric <- factor(plot_data$metric, 
                           levels = c("BPC/TIC Ratio", "Data Point Coverage"))

ggplot(plot_data, aes(x = rt_bin, y = value, color = lab, group = sample_id)) +
  geom_line(alpha = 0.7, linewidth = 1) +
  facet_grid(metric ~ mixture, scales = "free_y") +
  labs(
    title = "Chromatographic Fingerprint Comparison (Detected Data)",
    subtitle = "BPC/TIC Ratio and Data Point Coverage by Lab and Mixture",
    x = "Retention Time Bin",
    y = "Value (Ratio or Count)",
    color = "Laboratory"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12) 
  )

## for later similarity analysis i need to remove NA 
ratio_all[is.na(ratio_all)] <- 0
```

Need to cut or make much wider.

## Chromatogram  Similarity

### ratio along retention time. 

Inference from Detected Ratio DTW

This is a novel metric. It compares the "complexity fingerprint" of only the
detected signal. Clustering here would mean that labs have a similar pattern
of complexity for the peaks they detect across the gradient. This is a
high-level comparison of the quality of the detected features.

```{r, dtw-ratio, eval =eval_sim, fig.width=10, fig.height=10}
# --- 1. Load necessary libraries and data ---
try(library(pheatmap))
try(library(proxy))
try(library(dplyr))
try(library(tidyr))

# --- 3. Calculate the DTW similarity matrix [cite: 81] ---
# Transpose ratio_all so samples are rows
sim_matrix_med <- as.matrix(proxy::dist(t(ratio_all), 
                                        method = "dtw_lb",
                                        window.type = "sakoechiba",
                                        window.size = 5))
rownames(sim_matrix_med) <- colnames(ratio_all)
colnames(sim_matrix_med) <- colnames(ratio_all)

# --- 4. Create the Annotation Data Frame ---
annotation_df <- data.frame(
  Lab = tic_detect$lab,
  Mixture = tic_detect$mixture
)
rownames(annotation_df) <- colnames(ratio_all) # CRITICAL: names must match

# --- 5. Generate the Heatmap ---
pheatmap(
  sim_matrix_med,
  main = "DTW Similarity of BPC/TIC Ratio (Detected Signal)",
  
  # Add the annotation bars
  annotation_row = annotation_df,
  annotation_col = annotation_df,
  
  # Hide the unreadable text labels
  show_rownames = FALSE,
  show_colnames = FALSE,
  
  # Improve clustering and color
  clustering_distance_rows = as.dist(sim_matrix_med),
  clustering_distance_cols = as.dist(sim_matrix_med),
  clustering_method = "ward.D2",
  color = hcl.colors(50, "YlOrRd", rev = TRUE)
)
```

!!! adding a column of color by lab would be nice. 


```{r , rm-detect}
## delete the memory hungry object 
rm(tic_detect, bpc_detect, sp_full_detect)
```

# Annotated signal

Here we will ONLY keep the annotated signal.

```{r load-annotations}
## get this table from library generation
study_group <- "HE"
res_afekta <- read.csv(file.path("..", "afekta", "results",
                                 study_group, "ring_trial_library_HE.csv"))
res_hmgu <- read.csv(file.path("..", "hmgu", "results", 
                               study_group, "ring_trial_library_HE.csv"))
res_icl <- read.csv(file.path("..", "icl" , "results", 
                               study_group, "ring_trial_library_HE.csv"))
res_cembio <- read.csv(file.path("..", "cembio", "results",
                                 study_group, "ring_trial_library_HE.csv"))
```


```{r, an_signal_sp}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/sp_full_ann.RData")) {
  bpparam <- MulticoreParam(8)
  sp_afekta <- detect_signal(lab = "afekta", study_group = study_group,
                             annotated = TRUE, bpparam = bpparam)
  sp_hmgu <- detect_signal(lab = "hmgu", study_group = study_group, 
                           annotated = TRUE, bpparam = bpparam) 
  sp_icl <- detect_signal(lab = "icl", study_group = study_group, 
                          annotated = TRUE, bpparam = bpparam)
  sp_cembio <- detect_signal(lab = "cembio", study_group = study_group, 
                             annotated = TRUE, bpparam = bpparam)
  sp_full_ann <- c(sp_afekta, sp_hmgu, sp_icl, sp_cembio) 
  save(sp_full_ann, file = "object/sp_full_ann.RData")
} else {
  load("object/sp_full_ann.RData")
}
```

```{r, rm-ann}
rm(sp_afekta, sp_hmgu, sp_icl, sp_cembio)
```

Summarize the number of MS1 spectra per lab:

```{r nb-spectra-ann}
table(sp_full_ann$lab) 
```

Summarize the rtime range per lab (in seconds):

```{r rt-range-ann}
lapply(split(rtime(sp_full_ann), sp_full_ann$lab) , range)
```

## Basic metrics

### Base Peak Chromatogram (BPC) area

```{r load-bpc-ann}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/bpc_ann.RData")) {
  bpc_ann <- Chromatograms(sp_full_ann, summarize.method = "max",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab"))
  bpc_ann <- setBackend(bpc_ann, ChromBackendMemory())
  save(file = "object/bpc_ann.RData", bpc_ann)
} else {
  load("object/bpc_ann.RData")
}
```

We summarize the BPC area per mixture per lab by summing the intensity values.

Inference from Annotated BPC: 

This measures the sum of peak heights (apices) for only the annotated 
standards. This is a direct measure of quantitative sensitivity for the 
target compounds. The high variability shows that for the same compounds,
some labs get much stronger signals than others. This is a key finding for an
inter-lab comparison.

```{r sum-bpc-ann}
sbpc <- vapply(intensity(bpc_ann), sum, numeric(1))

plot_data_bpc_ann <- data.frame(
  Lab = bpc_ann$lab,
  Mixture = bpc_ann$mixture,
  BPC_Area = sbpc
)

ggplot(plot_data_bpc_ann, aes(x = Lab, y = log2(BPC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Annotated BPC Area Across Laboratories",
    subtitle = "Sum of annotated peak heights. Points colored by mixture type.",
    x = "Laboratory",
    y = "Log2(Annotated BPC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Next, we compute the coefficient of variation (CV) per mixture across labs.

```{r cv-ann}
cv_per_mix_ann <- tapply(sbpc, bpc_ann$mixture, function(x)
  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE))
plot(cv_per_mix_ann, main = "CV of Annotated BPC per Mixture")
```

### Total Ion Current (TIC) area

```{r load-tic-ann}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/tic_ann.RData")) {
  tic_ann <- Chromatograms(sp_full_ann, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab")) 
  tic_ann <- setBackend(tic_ann, ChromBackendMemory())
  save(file = "object/tic_ann.RData", tic_ann)
} else {
  load("object/tic_ann.RData")
}
```

We summarize the TIC area per mixture per lab by summing the intensity values.

This measures the sum of peak areas for only the annotated standards. This is
the best metric for overall quantitative agreement. The plots show 'afekta'
and cembio have 
the highest integrated area for these standards, while 'hmgu' and 'icl' have the lowest. 
The CV (calculated below) will be a key result for the thesis, quantifying 
the inter-lab quantitative discrepancy.

```{r int-tic-ann}
stic <- vapply(intensity(tic_ann), sum, numeric(1))

plot_data_tic_ann <- data.frame(
  Lab = tic_ann$lab,
  Mixture = tic_ann$mixture,
  TIC_Area = stic
)

ggplot(plot_data_tic_ann, aes(x = Lab, y = log2(TIC_Area), fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Annotated TIC Area Across Laboratories",
    subtitle = "Sum of annotated peak areas. Points colored by mixture type.",
    x = "Laboratory",
    y = "Log2(Annotated TIC Area)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### ratio BPC / TIC

Now we get the ratio for the spiked compound that were detected. 

Inference from Annotated BPC/TIC Ratio
This is a critical metric for peak purity. This ratio (Apex Height / Peak Area, 
scaled) for known compounds tells us how "clean" those peaks are.

A high ratio suggests a sharp, pure peak where the apex dominates the area.

A low ratio suggests a broader, tailing, or co-eluting peak, where the apex
intensity is low relative to the total peak area.

The plot shows 'afekta' and 'cembio' have the highest ratio (cleanest peaks), 
while 'icl' has the lowest (broadest/messiest peaks). This is a key difference
in chromatographic/MS performance.

```{r ratio-ann-across}
ratio_bpc_tic <- sbpc / stic

plot_data_ratio_ann <- data.frame(
  Lab = tic_ann$lab,
  Mixture = tic_ann$mixture,
  BPC_TIC_Ratio = ratio_bpc_tic
)

ggplot(plot_data_ratio_ann, aes(x = Lab, y = BPC_TIC_Ratio, fill = Lab)) +
  geom_violin(alpha = 0.5, trim = FALSE) +
  geom_quasirandom(aes(color = Mixture), dodge.width = 0.9, size = 2) +
  labs(
    title = "Annotated Data BPC/TIC Ratio Across Laboratories",
    subtitle = "Points colored by mixture type",
    x = "Laboratory",
    y = "BPC/TIC Ratio (Annotated Signal)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Max is one, why are the plot going above ? 

## tic/bpc ratio along rt 

Inference from Annotated BPC/TIC Ratio along RT
This plot shows where the annotated standards are eluting and what their peak 
purity (BPC/TIC ratio) is at that location.

BPC/TIC Ratio Plot: The plot should be "spiky," not continuous, as it only
contains data from annotated peaks. The height of the spikes in each bin 
reflects the average "purity" of the standards eluting there. We can see the 
'icl' profile is consistently lower than the others, confirming its peaks are
less "pure" (lower BPC/TIC ratio) across the entire gradient.

Data Point Coverage Plot: This is a key validation plot. It shows how many 
annotated standards were detected in each RT bin. It confirms that all labs 
are finding the standards in roughly the same elution regions (e.g., bins 5-15).

```{r ratio-ann, eval = FALSE, fig.width=12, fig.height=9}
# Preallocate containers
ratio_all <- sapply(seq_along(tic_ann), function(i) {
  rt <- rtime(tic_ann[i])[[1]]
  int_tic <- intensity(tic_ann[i])[[1]]
  int_bpc <- intensity(bpc_ann[i])[[1]]
  
  compute_bin_ratio(rt, int_tic, int_bpc, n_bins = 30)
})

coverage_all <- sapply(seq_along(tic_ann), function(i) {
  rt <- rtime(tic_ann[i])[[1]]
  compute_bin_count(rt, n_bins = 30)
})

# Assign names
colnames(ratio_all) <- paste(tic_ann$lab, tic_ann$mixture, "pol:", 
                             tic_ann$polarity, sep = "_")
colnames(coverage_all) <- colnames(ratio_all)

# --- 1. Melt the 'ratio_all' matrix ---
ratio_df <- as.data.frame(ratio_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "BPC/TIC Ratio") 

# --- 2. Melt the 'coverage_all' matrix ---
coverage_df <- as.data.frame(coverage_all) %>%
  mutate(rt_bin = 1:n()) %>%
  pivot_longer(
    cols = -rt_bin,
    names_to = "sample_id",
    values_to = "value"
  ) %>%
  mutate(metric = "Annotated Peak Count") # Renamed for clarity

# --- 3. Combine them and add lab/mixture info ---
sample_info <- data.frame(
  sample_id = colnames(ratio_all),
  lab = tic_ann$lab,
  mixture = tic_ann$mixture
)

plot_data <- bind_rows(ratio_df, coverage_df) %>%
  left_join(sample_info, by = "sample_id")

# --- 4. Plot ---
plot_data$metric <- factor(plot_data$metric, 
                           levels = c("BPC/TIC Ratio", "Annotated Peak Count"))

ggplot(plot_data, aes(x = rt_bin, y = value, color = lab, group = sample_id)) +
  geom_line(alpha = 0.7, linewidth = 1) +
  facet_grid(metric ~ mixture, scales = "free_y") +
  labs(
    title = "Chromatographic Fingerprint Comparison (Annotated Data)",
    subtitle = "BPC/TIC Ratio and Peak Count by Lab and Mixture",
    x = "Retention Time Bin",
    y = "Value (Ratio or Count)",
    color = "Laboratory"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 12) 
  )

## for later similarity analysis i need to remove NA 
ratio_all[is.na(ratio_all)] <- 0
```

Note: ye we will probably get rid of this, it's not readable anymore.

## Chromatogram  Similarity

This compares the "purity fingerprint" (from the binned BPC/TIC ratio plot)
between labs. Labs that cluster here have a similar pattern of peak purity
across the gradient for the annotated standards.

### ratio along retention time.

```{r ratio-sim , eval=eval_sim}
sim_matrix_med <- as.matrix(proxy::dist(t(ratio_all), 
                                        method = "dtw_lb",
                                        window.type = "sakoechiba",
                                        window.size = 5))

labels <- colnames(ratio_all)
rownames(sim_matrix_med) <- labels
colnames(sim_matrix_med) <- labels

# clustering
hc_med <- hclust(as.dist(sim_matrix_med))
plot(hc_med, labels = labels, main = "DTW clustering of BPC/TIC ratio (Annotated)")

# heatmap
heatmap(sim_matrix_med, labRow = labels, labCol = labels, 
        main = "DTW similarity of BPC/TIC ratio (Annotated)")
```

Plot summary plots:

```{r}
png("figure/tic_comparison_plot_pos.png", width = 10, height = 6,
    units = "in", res = 300)
plot_signal_comparison("TIC", 1)
dev.off()

png("figure/bpc_comparison_plot_pos.png", width = 10, height = 6,
    units = "in", res = 300)
plot_signal_comparison("BPC", 1)
dev.off()

png("figure/tic_comparison_plot_neg.png", width = 10, height = 6,
    units = "in", res = 300)
plot_signal_comparison("TIC", 0)
dev.off()

png("figure/bpc_comparison_plot_neg.png", width = 10, height = 6,
    units = "in", res = 300)
plot_signal_comparison("BPC", 0)
dev.off()
```


# Background/other signal

Here i need to load previous object and generate summary of "noise"

-   total signal - all detected peak signal

This value (dif_int2) represents the total baseline signal
(chemical and electronic noise) that was excluded by the peak picking
algorithm. A lab with a high value here (like 'afekta') has a much "noisier"
baseline that was successfully removed.

```{r noise-all}
load("object/tic_full.RData")
load("object/tic_detect.RData")

stic_full <- vapply(intensity(tic_full), sum, numeric(1))
stic_detect <- vapply(intensity(tic_detect), sum, numeric(1))

bg <- stic_full - stic_detect
ratio <- stic_detect/ bg

ggplot(data.frame(Lab = tic_full$lab, Ratio =log2(ratio)), aes(x = Lab, y = Ratio, fill = Lab)
) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  labs(
    title = "Detected vs Background TIC Area Ratio Across Laboratories",
    x = "Laboratory",
    y = "Detected / Background TIC Area Rati (log2)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

-   total signal - annotated peak signal

This value (dif_int3) represents all signal except the annotated standards. 
This includes the baseline noise and all the unannotated peaks.

```{r noise-ann}
load("object/tic_ann.RData")
stic_ann <- vapply(intensity(tic_ann), sum, numeric(1))
dif_int3 <- stic_full - stic_ann
ratio <- stic_ann/ dif_int3
ggplot(data.frame(Lab = tic_full$lab, Ratio =log2(ratio)), aes(x = Lab, y = Ratio, fill = Lab)
) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  labs(
    title = "Annotated vs Other TIC Area Ratio Across Laboratories",
    x = "Laboratory",
    y = "Annotated / Other TIC Area Ratio (log2)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

- detected - annotated peak signal

This (dif_int4) is a very insightful metric for my thesis. It represents the 
total signal from unannotated peaks. This is the "unknown" but "real" chemical
signal that was detected.

A lab with a high value (like 'icl') is detecting more "other" compounds. 
This could be due to:
Higher sensitivity, allowing it to see more low-level unknown compounds.
Different fragmentation/adduct formation, meaning it's detecting more variants 
of the annotated standards, which are not themselves annotated.

A lab with a low value (like 'hmgu') is detecting less signal outside of the 
annotated standards.

```{r noise-detect-ann}
ratio <- stic_ann/ (stic_detect - stic_ann)
ggplot(data.frame(Lab = tic_full$lab, Ratio =log2(ratio)),
       aes(x = Lab, y = Ratio, fill = Lab)
) +
  geom_violin(alpha = 0.5, trim = TRUE) +
  labs(
    title = "Annotated vs Other Detected TIC Area Ratio Across Laboratories",
    x = "Laboratory",
    y = "Annotated / Other Detected TIC Area Ratio (log2)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


# Peak analysis 

This section implements the downstream peak feature analysis you planned . 
We load the extracted EICs, clean them, calculate metrics, and then perform
inter-lab comparisons on a per-compound basis.

```{r load-tic-full-spectra}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/tic_full_spectra.RData")) {
  load("object/sp_full.RData")
  sp_full <- filterEmptySpectra(sp_full)
  tic_full <- Chromatograms(sp_full, summarize.method = "sum",
                       factorize.by= "dataOrigin",
                       spectraVariables = c("mixture", "lab")) 
  save(tic_full, file = "object/tic_full_spectra.RData") 
} else {  
  load("object/tic_full_spectra.RData")
}
```

## Prep raw data 

I think I will extract from TIC full data. I want to take some extra data as 
"padding" around the detected peaks. 

### extract peaks 

```{r peak-table}
res_afekta$lab <- "afekta"
res_hmgu$lab <- "hmgu"
res_icl$lab <- "icl"
res_cembio$lab <- "cembio"

peak_table <- rbind(res_hmgu, res_afekta, res_icl, res_cembio)

colnames(peak_table)[1] <- "chrom_peak_id"
colnames(peak_table)[4:7] <- c("mzMin", "mzMax", "rtMin", "rtMax")
peak_table$mzMin <- peak_table$mzMin - 0.003
peak_table$mzMax <- peak_table$mzMax + 0.003
peak_table$rtMin <- peak_table$rtMin - 3
peak_table$rtMax <- peak_table$rtMax + 3
peak_table$polarity <- ifelse(peak_table$polarity == "pos", 1, 0)
```


```{r load-eics-detected, warning = FALSE}
#| code-fold: true
#| code-summary: "Show the code"
if (!file.exists("object/eics_detected.RData")) {
  eics <- chromExtract(tic_full, peak.table = peak_table,
                       by = c("polarity", "lab", "mixture"))
  eics <- setBackend(eics, ChromBackendMemory())
  eics <- imputePeaksData(eics, method = "linear")
  eics <- applyProcessing(eics)
} else {
  load("object/eics_detected.RData")
}
```

-   number of peak per lab/ annotated adducts and annotated compound.

```{r peak-summary}
## number of peak per lab 
table(peak_table$lab)

## number of compound per lab
split(peak_table$compound_name, peak_table$lab) |> 
  lapply(unique) |> 
  lapply(length)
```

How may where there in HE ? 

```{r peak-lengths}
split(lengths(eics), eics$lab) |>
  lapply(summary)
```

I want to recalculate this after cleanup.

## Clean eics

```{r}
# --- 1. Define processing parameters ---
baseline <- BaselineParam(lambda = 200, p = 0.005)
smooth_params <- SavitzkyGolayParam(window = 9, polynomial = 2)
isolate_params <- IsolatePeakParam(frac = 0.05, tail_point = 1)

# --- 2. Apply the processing steps in a pipeline ---
# The pipe makes the sequence of operations very clear
final_eics <- eics |>
  processEICs(param = baseline) |>
  processEICs(param = smooth_params) |>
  processEICs(param = isolate_params) 

## get rid of empty eics (from isolation or bad extraction)
idx_na <- which(sapply(intensity(final_eics), function(i) all(is.na(i))) | 
                  lengths(intensity(final_eics)) < 3)

if (length(idx_na) > 0) {
  print(paste("Removing", length(idx_na), "EICs with all NA intensities or too short"))
  final_eics <- final_eics[-idx_na]
}
```

## Peak feature calculation

```{r}
# Calculate metrics using your custom function
metrics <- calculatePeakMetrics(peaksData(final_eics))

# Merge metrics with EIC metadata (lab, mixture, compound)
eic_data <- as.data.frame(chromData(final_eics))
full_peak_report <- cbind(eic_data, metrics)

# Filter out low-intensity peaks
idx_low_int <- which(full_peak_report$apex_intensity < 500)
if (length(idx_low_int) > 0) {
  warning(paste("Removing", length(idx_low_int), "peaks with apex intensity < 500"))
  full_peak_report <- full_peak_report[-idx_low_int, ]
}

head(full_peak_report)
```

Below i explore possibly more advanced filtering based on peak metrics. !!!
- base to apex ratio could help 
- check why some do nto have their FWHM calculated

```{r}
which(new_metrics$bs_ratio > 0.24) -> ids_test
dir.create("figure/peak_check/", showWarnings = FALSE)
for (i in ids_test) {
  png(paste0("figure/peak_check/eic_bad_", i, ".png"))
  plotChromatograms(final_eics[i], main = paste("EIC", i, "Bad"))
  dev.off()
}
```

- Some look of some feature specifically acccross lab (summary()) could be interesting

```{r}
# Entropy (from your calculatePeakMetrics function)
split(full_peak_report$entropy, full_peak_report$lab) |>
  lapply(summary)
```
Entropy is a measure of peak complexity or "purity".

A low entropy is good. It signifies a "pure," sharp, and simple peak shape where the signal is highly concentrated.

A high entropy is bad. It indicates a "messy" peak that is broad, noisy, or has significant tailing/fronting, as the signal is "spread out."
If one lab (e.t., 'icl') has a systematically higher median entropy than the others ('afekta', 'cembio'), it provides quantitative proof that their chromatographic setup produces less "pure" or less ideal peak shapes, which can negatively impact quantification.

```{r}
# FWHM (from your calculatePeakMetrics function)
split(full_peak_report$fwhm, full_peak_report$lab) |>
  lapply(summary)
```

FWHM is the classic measure of peak width.

A low FWHM is good. It means peaks are sharp and narrow, indicating high chromatographic efficiency and good separation.

A high FWHM is bad. It means peaks are broad, which leads to poorer resolution (more co-elution) and lower sensitivity (as the signal is spread over more time).


What to look for: This directly assesses the "Peak Shape Analysis" you planned. If 'hmgu' shows a significantly larger median FWHM, it suggests their LC system (or MS settings) is less efficient at producing sharp peaks than the other labs.

```{r}
# Gaussian Peak Fit R^2 (from your calculatePeakMetrics function)
split(full_peak_report$gaussian_similarity, full_peak_report$lab) |>
  lapply(summary)
```

```{r}
## tailing 
split(full_peak_report$tailing_factor, full_peak_report$lab) |>
  lapply(summary)
```
Tailing Factor measures peak symmetry.
A Tailing Factor close to 1 is ideal, indicating a symmetric peak shape.
A Tailing Factor significantly greater than 1 indicates peak tailing, where the trailing edge of the peak is longer than the leading edge. This is undesirable as it can lead to co-elution and inaccurate quantification.

A Tailing Factor significantly less than 1 indicates fronting, where the leading edge of the peak is longer than the trailing edge. This is also undesirable for similar reasons.


```{r}
## tailing 
split(full_peak_report$tailing_factor, full_peak_report$lab) |>
  lapply(summary)
```

## Downstream Peak Feature Analysis 

1. Retention Time (RT) Shift Analysis

Inference: RT Stability
This analysis  calculates the mean, standard deviation (SD), and coefficient 
of variation (CV) of the retention time for each compound across all labs.

Low CV: Indicates good chromatographic stability and high agreement between 
labs, as expected from a harmonized protocol.

High CV: Indicates significant RT shifts for that specific compound, 
suggesting it is sensitive to minor differences in LC setup (e.g., column 
temperature, gradient mixing) between labs.

```{r}
# We need to get the RT of the apex from the metrics
rt_report <- full_peak_report %>%
  group_by(compound_name, mixture) %>%
  summarize(
    mean_rt = mean(rt_apex, na.rm = TRUE),
    sd_rt = sd(rt_apex, na.rm = TRUE),
    cv_rt = sd_rt / mean_rt,
    n_detect = n()
  ) %>%
  arrange(desc(cv_rt))

print("Retention Time Stability Report (Top 10 most variable):")
head(rt_report, 10)

# Plot the distribution of RT CVs
ggplot(rt_report, aes(x = cv_rt)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.8) +
  labs(title = "Distribution of Retention Time CVs",
       x = "CV of RT (per compound)",
       y = "Frequency") +
  theme_minimal()
```


2. EIC Intensity Consistency (Quantitative Robustness)

Since the LC setup is harmonized and the same standard mixture is used, the 
total mass/amount injected is the same. Differences in measured peak 
area/height are due to variations in the MS (e.g., ionization efficiency, 
instrument tuning, mass accuracy/resolution, MS type).
Computation: Calculate the $\text{CV}$ of the Integrated Peak Area and Peak 
Height for each standard across the labs.Analysis: This is a crucial metric 
for ring trials, indicating the Quantitative Robustness of the MS methods.
High $\text{CV}$ suggests large differences in MS sensitivity.

Inference: Quantitative Robustness
This is a critical metric for a ring trial. It calculates the CV of peak Area 
(AUC) and Height (apex_intensity) for each compound across all labs.

Low CV: Indicates good quantitative agreement. The MS setups, despite being 
different, are providing similar quantitative responses for that compound.

High CV: This is a major finding. It indicates large differences in MS 
sensitivity  for that specific compound. This highlights which compounds are 
"hard to quantify" consistently and which labs' setups are outliers


```{r}
intensity_report <- full_peak_report %>%
  group_by(compound_name, mixture) %>%
  summarize(
    mean_auc = mean(auc, na.rm = TRUE),
    sd_auc = sd(auc, na.rm = TRUE),
    cv_auc = sd_auc / mean_auc,
    mean_height = mean(apex_intensity, na.rm = TRUE),
    sd_height = sd(apex_intensity, na.rm = TRUE),
    cv_height = sd_height / mean_height,
    n_detect = n()
  ) %>%
  arrange(desc(cv_auc))

print("Quantitative Robustness Report (Top 10 most variable by Area):")
head(intensity_report, 10)

# Plot the distribution of Area CVs
ggplot(intensity_report, aes(x = cv_auc)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.8) +
  labs(title = "Distribution of Peak Area (AUC) CVs",
       x = "CV of AUC (per compound)",
       y = "Frequency") +
  theme_minimal()
```

3. Principal Component Analysis (PCA) on Peak Features

Inference: Clustering by Lab or Compound
This analysis  clusters all detected peaks based on their shape and 
intensity features (RT, Area, Height, FWHM, etc.).

Strong clustering by Lab: This is a powerful conclusion. It means the
MS setup is the dominant factor driving peak quality. For example, all
peaks from 'icl' might cluster together because they are systematically
broader (low BPC/TIC ratio, as seen earlier) or have higher intensity.

Clustering by Compound: This would suggest that the labs are relatively
consistent, and the main differences in peak shape are driven by the
chemical properties of the compounds themselves.

No clear clustering: This indicates high variability (a "hairball" plot),
where no single factor dominates.

```{r, eval =FALSE}
# Prepare the data for PCA
# We need a numeric matrix: one row per peak, one col per feature
pca_data <- full_peak_report %>%
  select(rt_apex, apex_intensity, fwhm, rt_width, auc, entropy) %>%
  # Remove any rows with NAs (e.g., from failed FWHM calculation)
  na.omit()

## get idx of the rows kept after na.omit to match metadata
idx <- which(complete.cases(full_peak_report[, c("rt_apex", "apex_intensity", 
                                              "fwhm", "rt_width", "auc", "entropy")]))


# Log-transform intensity/area to manage scale differences
pca_data$apex_intensity <- log10(pca_data$apex_intensity)
pca_data$auc <- log10(pca_data$auc)

# Run PCA
pca_res <- prcomp(pca_data, center = TRUE, scale. = TRUE)

# Get the plotting data (scores)
pca_scores <- as.data.frame(pca_res$x)

# Add the metadata back
# We need to use the same 'na.omit()' logic to keep indices aligned
pca_scores$lab <- full_peak_report$lab[idx]
pca_scores$mixture <- full_peak_report$mixture[idx]
pca_scores$compound_name <- full_peak_report$compound_name[idx]

# Get variance explained
pc_var <- summary(pca_res)$importance[2,]

# Plot PC1 vs PC2
ggplot(pca_scores, aes(x = PC1, y = PC2, color = lab, shape = mixture)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "PCA of Peak Features",
    subtitle = "Do peaks cluster by lab, based on their shape and intensity?",
    x = paste0("PC1 (", round(pc_var[1] * 100, 1), "% variance)"),
    y = paste0("PC2 (", round(pc_var[2] * 100, 1), "% variance)"),
    color = "Laboratory",
    shape = "Mixture"
  ) +
  theme_minimal() +
  # Add ellipses to show the 95% confidence interval for each lab
  stat_ellipse(aes(group = lab), level = 0.95, linetype = "dashed")
```
 I need to fix the PCA plots.

Next step: 

Mass Accuracy Check (if possible from EIC data):

[probably in a separate file]
=
Using peak feature to do blidn clustering and see if this works as a way to group adducts together.
Some notes for this: 
- for similarity comparison try zeros grouping, check rt scan order. 
or difference between each data point. I think this could work. 







